---
title: "Using the `censored_cumulative` family"
author: "Martin Modrák"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Using the `censored_cumulative` family}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


The censored cumulative family extends the `cumulative(link = "logit")` family
to allow for observations that are only partially known. I.e. instead of observing
an integer, we observe a _range_ of integers that we cannot distinguish. 
Since ordinal response is bounded by definition, this lets us directly model also left and right censoring.

Mathematically, if we have $K$ categories, then given a set of thresholds $\tau_0 = -\infty < \tau_1 < ... < \tau_{K - 1} < \tau_K = \infty$
and observed bounds $l \leq u$
we have:

$$
\mathrm{Pr}(l \leq Y \leq u) = F(\tau_u - \mu) - F(\tau_{l - 1} - \mu)
$$
where $F$ is the CDF of the logistic distribution and $\mu$ is the linear predictor.

As a preprocessing step to using this family you need to make sure that all categories are actually distinguishable. 
I.e. if two categories are only ever observed in ranges that contain them both, they need to be merged before using the model.


```{r setup, message=FALSE,warning=FALSE, results="hide"}
library(brms)
library(brmsforce)
options(brms.backend = "cmdstanr", mc.cores = parallel::detectCores())

fit_cache_dir <- "fit_cache"
if(!dir.exists(fit_cache_dir)) {
  dir.create(fit_cache_dir)
}
```


Here's how we can simulate a simple dataset with 4 categories, matching the model (ignoring priors),
first without any censoring:


```{r}
set.seed(268523)
N <- 200
thresholds <- sort(rnorm(3))#c(-1, -0.3, 1.2)
main_effect <- 0.5
X <- rnorm(N)

#mu <- X * main_effect
mu <- 0
mu_noise <- mu + rlogis(N)
Y_no_cens <- dplyr::case_when(
  mu_noise < thresholds[1] ~ 1,
  mu_noise < thresholds[2] ~ 2,
  mu_noise < thresholds[3] ~ 3,
  TRUE ~ 4
)

if(length(unique(Y_no_cens)) != 4) {
  stop("Some categories not observed")
}


```

Now we can add some simple left and right censoring where for some
observations, values

```{r}
set.seed(965423)
bounds_raw <- matrix(sample(1:4, size = 2*N, replace = TRUE), ncol = 2, nrow = N)
low_bounds <- pmin(bounds_raw[,1], bounds_raw[,2])
high_bounds <- pmax(bounds_raw[,1], bounds_raw[,2])
#low_bounds <- sample(1:3, size = N, replace = TRUE)

Y_min <- Y_no_cens
Y_max <- Y_no_cens

left_cens <- Y_no_cens <= low_bounds
right_cens <- Y_no_cens >= high_bounds
Y_min[left_cens] <- 1
Y_max[left_cens] <- low_bounds[left_cens]

Y_min[right_cens] <- high_bounds[right_cens]
Y_max[right_cens] <- 4

df <- data.frame(Y_no_cens, Y_min, Y_max, X)
```

```{r}
hist(Y_max - Y_min, breaks = 0:4 - 0.5)
```


Now, let's fit the model:

```{r}
thres_prior <- prior(normal(0, 1), class = "Intercept")
fit <- brm(brmsformula(Y_min | vint(Y_max)  ~ 1, center = FALSE) , data = df, 
           family = censored_cumulative_logit(), stanvars = censored_cumulative_logit_stanvars(), 
           prior = thres_prior,
           file = file.path(fit_cache_dir, "censored_cumulative"), file_refit = "on_change",
           refresh = 500)

summary(fit)
```

From our simulation, the `Intercept[x]` correspond to the `thresholds` variable (`r paste0(thresholds, collapse = ", ")`) and `X` to the `main_effect` (`r main_effect`) so this lines up reasonably well.

```{r}
fit_no_cens <- brm(brmsformula(Y_no_cens ~ 1, center = FALSE) , data = df, 
           family = cumulative("logit"), 
           refresh = 500,
           prior = thres_prior,
           file = file.path(fit_cache_dir, "censored_cumulative_nocens"), file_refit = "on_change")
summary(fit_no_cens)

```

